---
title: "Multivariate time series: maximum likelihood estimation (Expectation-Maximization)"
author: "Javier Cara"
date: "2020-02"
output: 
  html_document:
    number_sections: true
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data

- Global temperature series from 1880 to 2015:
    - **globaltemp**: global mean land-ocean temperature index data.
    - **globaltempl**: surface air temperature index data using only meteorological satation data.

```{r}
load("globtemp.rda")
load("globtempl.rda")
y1 = globtemp
y2 = globtempl

# number of years to be predicted
num_pred = 2015-2006+1

# Multivariate data:
y = rbind(y1[1:(length(y1)-num_pred)],y2[1:(length(y1)-num_pred)])
ny = nrow(y)
nt = ncol(y)

# for comparison, data to be predicted
ya = rbind(y1[(nt+1):(nt+num_pred)],y2[(nt+1):(nt+num_pred)])
```

# Model

$$
\begin{bmatrix} 
y_{1,t} \\
y_{2,t}
\end{bmatrix}
= 
\begin{bmatrix} 
c_{11} & c_{12} & c_{13} & c_{14} \\
c_{21} & c_{22} & c_{23} & c_{24}
\end{bmatrix}
\begin{bmatrix}
x_{1,t} \\
x_{2,t} \\
x_{3,t} \\
x_{4,t}
\end{bmatrix}
+ 
\begin{bmatrix} 
v_{1,t} \\
v_{2,t}
\end{bmatrix}
$$

$$
v_t =
\begin{bmatrix} 
v_{1,t} \\
v_{2,t}
\end{bmatrix}
\sim 
N \left(  
\begin{bmatrix} 
0 \\
0
\end{bmatrix}
,
\begin{bmatrix} 
r_{11} & r_{12} \\
r_{12} & r_{22}
\end{bmatrix}
\right)
$$


$$
\begin{bmatrix}
x_{1,t+1} \\
x_{2,t+1} \\
x_{3,t+1} \\
x_{4,t+1}
\end{bmatrix}
=
\begin{bmatrix} 
a_{11} & a_{12} & a_{13} & a_{14} \\
a_{21} & a_{22} & a_{23} & a_{24} \\
a_{31} & a_{32} & a_{33} & a_{34} \\
a_{41} & a_{42} & a_{43} & a_{44} 
\end{bmatrix}
\begin{bmatrix}
x_{1,t} \\
x_{2,t} \\
x_{3,t} \\
x_{4,t}
\end{bmatrix}
+
\begin{bmatrix}
w_{1,t} \\
w_{2,t} \\
w_{3,t} \\
w_{4,t}
\end{bmatrix}
$$

$$
w_t =
\begin{bmatrix}
w_{1,t} \\
w_{2,t} \\
w_{3,t} \\
w_{4,t}
\end{bmatrix}
\sim N 
\left(  
\begin{bmatrix} 
0 \\
0 \\
0 \\
0
\end{bmatrix}
,
\begin{bmatrix} 
q_{11} & q_{12} & q_{13} & q_{14} \\
q_{21} & q_{22} & q_{23} & q_{24} \\
q_{31} & q_{32} & q_{33} & q_{34} \\
q_{41} & q_{42} & q_{43} & q_{44} 
\end{bmatrix}
\right)
$$

$$
x_1 
=
\begin{bmatrix} 
x_{1,1} \\
x_{1,2} \\
x_{1,3} \\
x_{1,4}
\end{bmatrix}
\sim 
N \left(  
\begin{bmatrix} 
m_{1,1} \\
m_{1,2} \\
m_{1,3} \\
m_{1,4}
\end{bmatrix}
,
\begin{bmatrix} 
p_{1,11} & p_{1,12} & p_{1,13} & p_{1,14} \\
p_{1,21} & p_{1,22} & p_{1,23} & p_{1,24} \\
p_{1,31} & p_{1,32} & p_{1,33} & p_{1,34} \\
p_{1,41} & p_{1,42} & p_{1,43} & p_{1,44} 
\end{bmatrix}
\right)
$$

- In total, 64 parameters!

- The number of states is

```{r}
nx = 4
```

# MLE using EM algorithm

## Starting values: subspace algorithm

```{r}
ssm_sub = emssm::ACQRS_sub(y,nx,ny)
Ai = ssm_sub$A
Ci = ssm_sub$C
Qi = ssm_sub$Q
Ri = ssm_sub$R
m1i = rep(0,nx)
P1i = diag(nx)
```

## EM algorithm

```{r}
ssm_e = emssm::ACQR_em(y,Ai,Ci,Qi,Ri,m1i,P1i,max_iter = 20,ny = ny,nx = nx,txo = T)
```

- The estimates of the parameters are:

```{r}
(Ae = ssm_e$A)
```

```{r}
(Qe = ssm_e$Q)
```

```{r}
(Ce = ssm_e$C)
```

```{r}
(Re = ssm_e$R)
```

```{r}
(m1e = ssm_e$m1)
```

```{r}
(P1e = ssm_e$P1)
```

# Estimated time series

```{r}
kf <- emssm::ACQR_kfilter(y,Ae,Ce,Qe,Re,m1e,P1e,nx,ny)
ye <- Ce %*% kf$xtt1[,-nt]
```

- Standar error of the estimated values

```{r}
ye_se = matrix(0, nrow = ny, ncol = nt)
for (t in 1:nt){
  ye_var = Ce %*% kf$Ptt1[,,t] %*% t(Ce) + Re # variance of the estimated values
  ye_se[,t] = sqrt(diag(ye_var))
}
```

- Confidence intervals of the estimated values

```{r}
alpha = 0.05
ye1 = ye - qnorm(1-alpha/2)*ye_se
ye2 = ye + qnorm(1-alpha/2)*ye_se
```

- Plots

```{r}
plot(y[1,], type="l", ylim = c(-1,1), ylab = "land-ocean index", xlab = "Time", col = 4)
lines(ye[1,], col = "gray", lwd = 2)
legend("topleft", legend = c("measured data","estimated data"), col = c("blue","gray"), lty = 1)
tt = 1:nt
xx = c(tt, rev(tt))
yy = c(ye1[1,], rev(ye2[1,]))
polygon(xx,yy, border = NA, col = gray(0.7, alpha = 0.3))
```

```{r}
plot(y[2,], type="l", ylim = c(-1,1), ylab = "land-only index", xlab = "Time", col = 3)
lines(ye[2,], col = "gray", lwd = 2)
legend("topleft", legend = c("measured data","estimated data"), col = c("green","gray"), lty = 1)
tt = 1:nt
xx = c(tt, rev(tt))
yy = c(ye1[2,], rev(ye2[2,]))
polygon(xx,yy, border = NA, col = gray(0.7, alpha = 0.3))
```

- Mean square errors

```{r}
apply((y-ye)^2,1,mean)
```


# Checking the residuals

```{r}
res <- y - ye
```

```{r}
plot(res[1,], type = 'l', ylab = "Residuals of land-ocean index", col = 4)
```

```{r}
plot(res[2,], type = 'l', ylab = "Residuals of land-only index", col = 3)
```

- The residuals must be white noise

```{r}
acf(res[1,])
```

```{r}
acf(res[2,])
```

# Predicting new values

## Year by year

```{r}
yp = matrix(0, nrow = ny, ncol = num_pred)
ypi1 = matrix(0, nrow = ny, ncol = num_pred)
ypi2 = matrix(0, nrow = ny, ncol = num_pred)
for (t in 1:num_pred){
  y_upd = cbind(y,ya[,t-1]) # updated data
  pred = emssm::ACQR_predict(y_upd,Ae,Ce,Qe,Re,m1e,P1e,nx,ny,n_ahead=1)
  yp[,t] = pred$yp
  ypi1[,t] = pred$ypi1
  ypi2[,t] = pred$ypi2
}
```


```{r}
plot(y[1,], type="l", ylim = c(-1,1), xlim = c(0,140), ylab = "land-ocean index", xlab = "Time", col = 4)
lines((nt+1):(nt+num_pred),ya[1,], col = 4, lty = 3, lwd = 2)
lines((nt+1):(nt+num_pred), yp[1,], col = "gray", lwd = 2)
tt = (nt+1):(nt+num_pred)
xx = c(tt, rev(tt))
yy = c(ypi1[1,], rev(ypi2[1,]))
polygon(xx,yy, border = NA, col = gray(0.7, alpha = 0.3))
```

```{r}
plot(y[2,], type="l", ylim = c(-1,1), xlim = c(0,140), ylab = "land-only index", xlab = "Time", col = 3)
lines((nt+1):(nt+num_pred),ya[2,], col = 3, lty = 3, lwd = 2)
lines((nt+1):(nt+num_pred), yp[2,], col = "gray", lwd = 2)
tt = (nt+1):(nt+num_pred)
xx = c(tt, rev(tt))
yy = c(ypi1[2,], rev(ypi2[2,]))
polygon(xx,yy, border = NA, col = gray(0.7, alpha = 0.3))
```

- Prediction errors:

```{r}
apply((ya-yp)^2,1,mean)
```

## Long term prediction

```{r}
pred = emssm::ACQR_predict(y,Ae,Ce,Qe,Re,m1e,P1e,nx,ny,n_ahead=num_pred)
```


```{r}
plot(y[1,], type="l", ylim = c(-1,1), xlim = c(0,140), ylab = "land-ocean index", xlab = "Time", col = 4)
lines((nt+1):(nt+num_pred),ya[1,], col = 4, lty = 3, lwd = 2)
lines((nt+1):(nt+num_pred), pred$yp[1,], col = "gray", lwd = 2)
tt = (nt+1):(nt+num_pred)
xx = c(tt, rev(tt))
yy = c(pred$ypi1[1,], rev(pred$ypi2[1,]))
polygon(xx,yy, border = NA, col = gray(0.7, alpha = 0.3))
```

```{r}
plot(y[2,], type="l", ylim = c(-1,1), xlim = c(0,140), ylab = "land-only index", xlab = "Time", col = 3)
lines((nt+1):(nt+num_pred),ya[2,], col = 3, lty = 3, lwd = 2)
lines((nt+1):(nt+num_pred), pred$yp[2,], col = "gray", lwd = 2)
tt = (nt+1):(nt+num_pred)
xx = c(tt, rev(tt))
yy = c(pred$ypi1[2,], rev(pred$ypi2[2,]))
polygon(xx,yy, border = NA, col = gray(0.7, alpha = 0.3))
```

- Prediction errors:

```{r}
apply((ya-pred$yp)^2,1,mean)
```
