---
title: "Global Warming: maximum likelihood estimation (Newton-Raphson)"
author: "Javier Cara"
date: "Nov 2018"
output: 
  html_document:
    number_sections: true
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data

- Global temperature series from 1880 to 2015:
    - **globaltemp**: global mean land-ocean temperature index data.
    - **globaltempl**: surface air temperature index data using only meteorological satation data.

```{r}
# more details in package 'astsa'
load("globtemp.rda")
load("globtempl.rda")
y1 = globtemp
y2 = globtempl
ts.plot(y1,y2, col = c(4,3))
legend("topleft", legend = c("land-ocean index","land-only index"), col = c(4,3), lty = 1)
```

- We are going to use data from 1880 to 2005 to estimate a state-space model. Then, 2006 to 2015 will be predicted using this model.

```{r}
# number of years to predict
num_pred = 2015-2006+1
```

- Multivariate data:

```{r}
y = rbind(y1[1:(length(y1)-num_pred)],y2[1:(length(y1)-num_pred)])
ny = nrow(y)
nt = ncol(y)
```

```{r}
# for comparison, data to be predicted
ya = rbind(y1[(nt+1):(nt+num_pred)],y2[(nt+1):(nt+num_pred)])
```

# Model

$$
\begin{bmatrix} 
y_{1,t} \\
y_{2,t}
\end{bmatrix}
= 
\begin{bmatrix} 
c_1 \\
c_2
\end{bmatrix}
x_t + 
\begin{bmatrix} 
v_{1,t} \\
v_{2,t}
\end{bmatrix}
, \quad v_t \sim N \left(  
\begin{bmatrix} 
0 \\
0
\end{bmatrix}
,
\begin{bmatrix} 
r_{11} & r_{12} \\
r_{12} & r_{22}
\end{bmatrix}
\right)
$$

$$
x_{t+1} = a x_{t} + w_t, \quad w_t \sim N(0,Q)
$$

$$
x_1 \sim N(m_1,P_1)
$$

- Parameters

$$
\theta = \{a,Q,c_1,c_2,r_{11},r_{12},r_{22},m_1,P_1 \}
$$

- The number of states is

```{r}
nx = 1
```

# MLE using Newton-Raphson

- Definition of the log-likelihood function:

```{r}
logLikelihood <- function(param){
  A = param[1]
  Q = param[2]^2
  C = matrix(c(param[3],param[4]),nrow=2)
  Rc = matrix(c(param[5],param[6],0,param[7]),nrow=2) # cholesky factorization
  R = Rc %*% t(Rc)
  m1 = param[8]
  P1 = param[9]^2
  kf = emssm::ACQR_kfilter(y,A,C,Q,R,m1,P1,nx,ny)
  return(-kf$loglik) # minus loglik because optim finds the minimum
}
```

- Starting values:

```{r}
# param = {phi,Q^{1/2},R^{1/2},m1,P1^{1/2}}
init_par = c(1,1,1,1,1,1,1,0,1)
```

- Estimation:

```{r}
ssm_e = optim(init_par,logLikelihood, gr = NULL, method = "BFGS", hessian = TRUE,
               control = list(trace=1, REPORT = 1, maxit = 20))
```

- The estimates of the parameters are:

```{r}
ssm_e$par
```

- Or in terms of the parameters of the model:

```{r}
(Ae = ssm_e$par[1])
```

```{r}
(Qe = ssm_e$par[2]^2)
```

```{r}
(Ce = matrix(c(ssm_e$par[3],ssm_e$par[4]), nrow = 2))
```

```{r}
Rc = matrix(c(ssm_e$par[5],ssm_e$par[6],0,ssm_e$par[7]),nrow=2) # Cholesky factor
(Re = Rc %*% t(Rc))
```

```{r}
(m1e = ssm_e$par[8])
```

```{r}
(P1e = ssm_e$par[9]^2)
```

# Estimated time series

```{r}
kf <- emssm::ACQR_kfilter(y,Ae,Ce,Qe,Re,m1e,P1e,nx,ny)
ye <- Ce %*% kf$xtt1[,-nt]
```

- Standar error of the estimated values

```{r}
ye_se = matrix(0, nrow = ny, ncol = nt)
for (t in 1:nt){
  ye_var = Ce %*% kf$Ptt1[,,t] %*% t(Ce) + Re # variance of the estimated values
  ye_se[,t] = sqrt(diag(ye_var))
}
```

- Confidence intervals of the estimated values

```{r}
alpha = 0.05
ye1 = ye - qnorm(1-alpha/2)*ye_se
ye2 = ye + qnorm(1-alpha/2)*ye_se
```

- Plots

```{r}
plot(y[1,], type="l", ylim = c(-1,1), ylab = "land-ocean index", xlab = "Time", col = 4)
lines(ye[1,], col = "gray", lwd = 2)
legend("topleft", legend = c("measured data","estimated data"), col = c("black","gray"), lty = 1)
tt = 1:nt
xx = c(tt, rev(tt))
yy = c(ye1[1,], rev(ye2[1,]))
polygon(xx,yy, border = NA, col = gray(0.7, alpha = 0.3))
```

```{r}
plot(y[2,], type="l", ylim = c(-1,1), ylab = "land-only index", xlab = "Time", col = 3)
lines(ye[2,], col = "gray", lwd = 2)
legend("topleft", legend = c("measured data","estimated data"), col = c("black","gray"), lty = 1)
tt = 1:nt
xx = c(tt, rev(tt))
yy = c(ye1[2,], rev(ye2[2,]))
polygon(xx,yy, border = NA, col = gray(0.7, alpha = 0.3))
```

- Mean square errors

```{r}
apply((y-ye)^2,1,mean)
```


# Checking the residuals

```{r}
res <- y - ye
```

```{r}
plot(res[1,], type = 'l', ylab = "Residuals of land-ocean index", col = 4)
```

```{r}
plot(res[2,], type = 'l', ylab = "Residuals of land-only index", col = 3)
```

- The residuals must be white noise

```{r}
acf(res[1,])
```

```{r}
acf(res[2,])
```

# Predicting new values

## Year by year

```{r}
yp = matrix(0, nrow = ny, ncol = num_pred)
ypi1 = matrix(0, nrow = ny, ncol = num_pred)
ypi2 = matrix(0, nrow = ny, ncol = num_pred)
for (t in 1:num_pred){
  y_upd = cbind(y,ya[,t-1]) # updated data
  pred = emssm::ACQR_predict(y_upd,Ae,Ce,Qe,Re,m1e,P1e,nx,ny,n_ahead=1)
  yp[,t] = pred$yp
  ypi1[,t] = pred$ypi1
  ypi2[,t] = pred$ypi2
}
```


```{r}
plot(y[1,], type="l", ylim = c(-1,1), xlim = c(0,140), ylab = "land-ocean index", xlab = "Time", col = 4)
lines((nt+1):(nt+num_pred),ya[1,], col = 4, lty = 3, lwd = 2)
lines((nt+1):(nt+num_pred), yp[1,], col = "gray", lwd = 2)
tt = (nt+1):(nt+num_pred)
xx = c(tt, rev(tt))
yy = c(ypi1[1,], rev(ypi2[1,]))
polygon(xx,yy, border = NA, col = gray(0.7, alpha = 0.3))
```

```{r}
plot(y[2,], type="l", ylim = c(-1,1), xlim = c(0,140), ylab = "land-only index", xlab = "Time", col = 3)
lines((nt+1):(nt+num_pred),ya[2,], col = 3, lty = 3, lwd = 2)
lines((nt+1):(nt+num_pred), yp[2,], col = "gray", lwd = 2)
tt = (nt+1):(nt+num_pred)
xx = c(tt, rev(tt))
yy = c(ypi1[2,], rev(ypi2[2,]))
polygon(xx,yy, border = NA, col = gray(0.7, alpha = 0.3))
```

- Prediction errors:

```{r}
apply((ya-yp)^2,1,mean)
```


## Long term prediction

```{r}
pred = emssm::ACQR_predict(y,Ae,Ce,Qe,Re,m1e,P1e,nx,ny,n_ahead=num_pred)
```


```{r}
plot(y[1,], type="l", ylim = c(-1,1), xlim = c(0,140), ylab = "land-ocean index", xlab = "Time", col = 4)
lines((nt+1):(nt+num_pred),ya[1,], col = 4, lty = 3, lwd = 2)
lines((nt+1):(nt+num_pred), pred$yp[1,], col = "gray", lwd = 2)
tt = (nt+1):(nt+num_pred)
xx = c(tt, rev(tt))
yy = c(pred$ypi1[1,], rev(pred$ypi2[1,]))
polygon(xx,yy, border = NA, col = gray(0.7, alpha = 0.3))
```

```{r}
plot(y[2,], type="l", ylim = c(-1,1), xlim = c(0,140), ylab = "land-only index", xlab = "Time", col = 3)
lines((nt+1):(nt+num_pred),ya[2,], col = 3, lty = 3, lwd = 2)
lines((nt+1):(nt+num_pred), pred$yp[2,], col = "gray", lwd = 2)
tt = (nt+1):(nt+num_pred)
xx = c(tt, rev(tt))
yy = c(pred$ypi1[2,], rev(pred$ypi2[2,]))
polygon(xx,yy, border = NA, col = gray(0.7, alpha = 0.3))
```

- Prediction errors:

```{r}
apply((ya-pred$yp)^2,1,mean)
```

